---
title: "Influence of Age and Past Vote Experience on Election results"
author: "Michelle Huang,  Chenxuan Ding, Yantong Li, Zuzhang Zhao"
date: '2020-10-05'
output: pdf_document
latex_engine: xelatex
---

```{r setup ,echo = FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

link to Github:


\newpage
# Executive summary

## Objective of the study

Petit Poll has been asked by the Liberal Party and a team consists of Michelle Huang, Chenxuan Ding, Yantong Li and Zuzhang Zhao to survey election pattern in Ontario for the upcoming Canadian Election. We would like to investigate how the diﬀerence in people’s attributes (age, previous voting etc.) could inﬂuence the election results.

## Impletemetation on the study

We surveyed 5000 people and recorded their answered surveyed data to analyze the data. We created a survey that would eventually go to the 5000 registered household owners living in Ontario in the government database using SurveyMonkey website. We asked various questions with some of the variables of interest.

## Discovery of the study

We found that the election results i.e. the party that each individual actually voted for can be greatly inﬂuenced by that individual’s age group and the past voting experience of whether they voted before or not. These two variables were included in the survey questionnaire and other questions in the survey questionnaire includes the current status in Canada and their personal interest on voting.

## Two examples 

We hypothesized that the diﬀerent people with diﬀerent status in Canada may have diﬀerent interests on voting, which result in diﬀerent vote results.

- For instance, permanent residents or Canadian citizen may pay much more attention to the election results compared that to group of people with study or work permit since the future due to the election results would have a much more signiﬁcant impact for permanent residents or Canadian citizen. Furthermore, the diﬀerence in political direction or opinions from each election candidates also have an inﬂuence on the mentioned variables.

- Another example is that if the election candidate cared very much about domestic elderly welfare, the permanent residents or Canadian citizen at an upper year group may vote for the party that this speciﬁc election candidate is in. Another example could be, one election candidate promotes to release the ﬁnancial burden for student loans for international students, it will become obvious the party of this election candidate would be voted by the group of people with younger year group that have current status with study permit.

## Meaning and use of the study

This survey could provide us with a great amount of insights on why people vote for certain party and let us to understand the pattern behind the scene by a data-driven procedure. This also allows us to provide the Liberal Party, in this case our client to understand the way how they could have a larger chance to be elected.

## Weakness and next step

The complicated interaction between each variable could introduce unreliability and uncertainty to our analysis, the next step could be rigorous modeling buildings and assumption checking of any correlation between each variable. And any lurking variables or confounding variables could be deeply investigated for any bias in our study.

\newpage
# Introduction

The question we are interested in is that what potential variables could aﬀect the Canadian election results and ﬁnd out how much the eﬀect could be, especially considering the voters for the Liberal Party. We created a survey that would eventually go to the 5000 registered household owners living in Ontario in the government database using SurveyMonkey website. We asked various questions with some of the variables of interest. We will collected the data and analyse them to draw the conclusion in the ﬁnal report.

## Survey methodology

Population frame and sample Population 
- All the household owners living in Ontario.

## Sampling frame 
- A list of 5000 registered household owners living in Ontario who answered the survey questionnaires.

## Sample 
- 5000 registered household owners people living in Ontario.

## Desired respondents acquisition

- Send the link of the web survey to the phone number to the registered household owners living in Ontario.

## Survey cost estimation

- $(15 + 0.1) * 5000 + ﬁx cost of survey design

- Starbucks gift card of $15 value

- Text fee $0.1 for sending the survey website link

- The total cost of the survey could be high due to the fact that we need to provide people completed the survey with incentive of gift cards, whereas the actual cost due to the survey logistics is not too much, only $0.1 of text fee

## Non-response and its effect on survey

- For each of response, give them starbucks gift card of $15 value. By doing this, we have some incentive for the people to do the survey. The group of people who are not quite interested in voting or do not care about election at all may be the non-response.

## Respondent privacy Protection

- Send the web link with some built-in privacy protocols which require ﬁlling in some personal information and conﬁrming the individual before the survey starts.

## Sampling methods and statistical properties

- I will use SRSWOR Sampling since this the most straightforward method and can avoid any potential variables causing bias in the results.

- I will not choose SRSWOR stratiﬁed or clustered at ﬁrst since we do not know whether each group or each individual could be representative for the whole population.

```{r}
library("tidyverse")
set.seed(500)

# Sample
# Current Status in Canada

status <- sample(c("Canadian citizen","Permanent resident","Study or work permit","Refugees"), size = 5000, prob = c(0.25,0.35,0.1,0.05), replace = TRUE)
interest <- sample(c("Very much","Not too much","Not at all"),size = 5000, prob = c(1/3,1/3,1/3), replace = TRUE)
age <- rnorm(5000, mean = 55, sd = 200)
vote_before <- sample(c("Yes","No"),size = 5000, prob = c(0.5,0.5), replace = TRUE)
# https://www.ctvnews.ca/canada/ontario-election-2018/ontario-election-results-since-1867-1.3962151
vote <- sample(c("Liberal Party","Conservative Party", "Others"),size = 5000, prob = c(0.58,0.28, 0.14), replace = TRUE)
vote_df <- tibble(status, interest, age,vote_before,vote)


# Population
status <- sample(c("Canadian citizen","Permanent resident","Study or work permit","Refugees"), size = 50000, prob = c(0.65,0.25,0.08,0.02), replace = TRUE)
interest <- sample(c("Very much","Not too much","Not at all"),size = 50000, prob = c(1/3,1/3,1/3), replace = TRUE)
age <- age <- rnorm(50000, mean = 55, sd = 200)
vote_before <- sample(c("Yes","No"),size = 50000, prob = c(0.5,0.5), replace = TRUE)
vote <- sample(c("Liberal Party","Conservative Party", "Others"),size = 50000, prob = c(0.58,0.28, 0.14), replace = TRUE)
vote_population_df <- tibble(status, interest, age,vote_before,vote)



# with some correlation
vote_before <- sample(c("Yes","No"),size = 50000, prob = c(0.5,0.5), replace = TRUE)

interest <- NULL
age <- NULL
status <- NULL
vote <- NULL

for (i in 1 : 50000){
  if (vote_before[i] == "No"){
    age[i] <- rnorm(1, mean = 25, sd = 5)
    if(age[i]  < 25){
      status[i] <- sample(c("Canadian citizen","Permanent resident","Study or work permit","Refugees"),1, prob = c(0.25,0.35,0.1,0.05), replace = TRUE)
      interest[i] <- sample(c("Very much","Not too much","Not at all"),1, prob = c(1/8,3/8,1/2), replace = TRUE)
      vote[i] <- sample(c("Liberal Party","Conservative Party", "Others"),1, prob = c(0.1,0.2, 0.7), replace = TRUE)
    }
    else{
      vote[i] <- sample(c("Liberal Party","Conservative Party", "Others"),1, prob = c(0.6,0.2, 0.2), replace = TRUE)
    }
  }
  if (vote_before[i] == "Yes"){
    age[i] <- rnorm(1, mean = 45, sd =8)
    vote[i] <- sample(c("Liberal Party","Conservative Party", "Others"),1, prob = c(0.3,0.4, 0.3), replace = TRUE)
    interest[i] <- sample(c("Very much","Not too much","Not at all"),1, prob = c(3/8,1/2,1/8), replace = TRUE)
    status[i] <- sample(c("Canadian citizen","Permanent resident","Study or work permit","Refugees"),1, prob = c(0.65,0.15,0.15,0.05), replace = TRUE)

  }
}

population <- tibble(status, interest, age,vote_before,vote)

```

\newpage
# Age vs Past Voting Experience

```{r}
p1 <- population %>%  ggplot(aes(x = vote_before, y = age)) + geom_boxplot() 
p1
```
We can clearly see that older people tends to have more voting experience before. So, we may be interested in how much of diﬀerence the age made on the past voting experience by calculating the mean and standard deviations of the two diﬀerence groups of with or without past voting experiences.
```{r message=FALSE}
sample_STRS <- population %>% group_by(vote_before) %>% sample_n(500)
sample_STRS %>% 
  group_by(vote_before) %>% 
  summarise(mean_age = mean(age), median_age = median(age))
```
We can see that people without any voting experience on average is around 25 whereas people with voting experience on average is around 45, there is 20 years old of diﬀerence.

\newpage


```{r}
p2 <- population %>%  ggplot(aes(x = vote, y = age, fill = vote_before)) + geom_boxplot()
p2
```

From this graph, we could see that the diﬀerence between past voting experience is much more signiﬁcant than the diﬀerent parties people vote. In other words, the group we use should be the past voting experience instead of what they vote for.


```{r}
set.seed(1)
sample_STRS <- population %>% group_by(vote_before) %>% sample_n(500)
sample_SRSWOR <- population %>% sample_n(1000)
```


```{r message=FALSE}
sample_STRS %>%  
  ggplot(aes(x = vote, fill = vote_before)) + geom_bar() + facet_wrap(~vote_before) + coord_flip()

```

From the ﬁrst vote distribution plot of votes for three diﬀerent parties voted, we could observe that for people without any voting experience vote others the most while people vote for Conservative Party the most for people with past voting experience.


\newpage

```{r}
sample_STRS %>%  
  ggplot(aes(x = vote, y = age)) + geom_boxplot()
sample_STRS %>% 
  group_by(vote) %>% 
  summarise(mean_age = mean(age), median_age = median(age))
```

From the second plot, we could see that people vote for Conservative Party are on average the older people whereas young people tend to vote for others maybe due to that they are not interested in voting.

# Results

From the analysis we performed, we determine that the past experience of voting may impact the ﬁnal results. The result of voting on the other hand may large impacted by the age groups as well since we found a quite large diﬀerences of results for diﬀerence age groups. This diﬀerence may be explained by the fact that the younger people are generally lacking of experience on voting in the past and may be not interesting in voting at all. On the other hand, older people tends to have much more experience on voting and would be interested in voting.

# Weakness and areas for future

There are some weakness including lacking of the source of data accuracy. Since we used the simulation based on the past census data from the website, the data could be outdated and some variables may lack of data scource at all. This results in inaccuracy of the simulation and not be able to recreate a real world scenario.


# References

- Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686

- R Core Team (2020). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.

- Government of Canada, S. (2019, June 18). Census Proﬁle, 2016 Census Canada [Country] and Canada [Country]. Retrieved October 05, 2020, from https://www12.statcan.gc.ca/census-recensement/2016/dp-pd/prof /details/Page.cfm?Lang=E

- Tahirali, J. (2018, June 07). Ontario election results since 1867. Retrieved October 05, 2020, from https://www.ctvnews.ca/canada/ ontario-election-2018/ontario-election-results-since-1867-1.3962151

- Hadley Wickham, Romain François, Lionel Henry and Kirill Müller (2020). dplyr: A Grammar of Data Manipulation. R package version 1.0.2. https://CRAN.R-project.org/package=dplyr

- H. Wickham. ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York, 2016.

- Yihui Xie (2020). knitr: A General-Purpose Package for Dynamic Report Generation in R. R package version 1.30.

- Yihui Xie (2015) Dynamic Documents with R and knitr. 2nd edition. Chapman and Hall/CRC. ISBN 978-1498716963

- Yihui Xie (2014) knitr: A Comprehensive Tool for Reproducible Research in R. In Victoria Stodden, Friedrich Leisch and Roger D. Peng, editors, Implementing Reproducible Computational Research. Chapman and Hall/CRC. ISBN 978-1466561595
